Glyph-maps for Visually Exploring Temporal Patterns in Climate Data and Models

### 2012-03-06 revision/paper.tex
done completely!

In both of the examples where locations have been collapsed, the icon locations still represent an actual geographic location of collected data, albeit including data from up to an icon-width away.  An alternative way to combine stations is to round locations to the icon size, as done in Figure \ref{fig:irregular-grid}. This produces a regular grid with the advantage of easier perception of structure \citet{Kr12b}. (Figure 9.12 of \citet{IPCC} uses a regular gridded time series over a spatial domain to illustrate temporal patterns in climate change data.) It hides the irregular nature of the locations, though, and a complication is that icons may end up centered over nonsense places, such as land measurements in the ocean.

@@

Glyphs can be more general than time series or stars. For multivariate data we may want to represent several variables in the glyphs. Figure~\ref{fig:cloud} shows scatterplots of the EXPO data as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled. The bivariate relationship between the two variables is explored spatially, and it is clear that it is different across the region. The equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association. The second plot displays a loess curve fit to the data instead of the individual points, sharpening the view of association. Another option is available in Gauguin \citep{gribov:2006}, a icon based on a histogram, giving small univariate distribution displays for each location.

@@

There are two natural next directions for this work: perceptual experiments, and development of interactive graphics. Several careful studies comparing the perceptual elements of these displays, using different scale and glyph options, and in comparison with facetted heatmaps, would be recommended. For example, it appears that periodic trends are easier to perceive in star-glyphs, and long-term trends in line-glyphs. Is this always the case? Are their certain types of periodicity that are particularly easy to spot? These questions need to be answered with rigorous perceptual studies to help guide the use of these plots.

The conceptualization and equations for computing the glyph-maps lend themselves to producing these displays interactively. Different variables can be swept into the display using mouse motion, resolution could be increased upon zooming in to a spatial neighborhood. This experimentation is ideally done using R, so that computation can be linked with displays. The hurdle is the lagging development of interactive graphics in R. Several new packages that support interactive graphics in R have become available in the past year \citep{qtbase, qtpaint, plumbr} which will enable these experiments.

In summary, glyph-maps enable the study of temporal patterns in multivariate spatiotemporal data. Climate change is focused on changes over time, so glyph-maps provide a way to study these changes directly. Glyph-maps enable different resolutions of the data to be examined: the raw data to discover data quality issues, global trend, seasonality, residuals, and multivariate dependencies. When the data is spatially gridded the glyphs are organized in a manner that makes reasonable comparisons. For irregularly gridded data the icon size needs to be chosen to minimize overlap but use as much display space as possible, or combined to produce close to gridded icons.

>>>>

In both of the examples where locations have been collapsed, the icon locations still represent an actual geographic location of collected data, albeit including data from up to an icon-width away.  An alternative way to combine stations is to round locations to the icon size, as done in Figure \ref{fig:irregular-grid}. This produces a regular grid with the advantage of easier perception of structure \citep{Kr12b}. (Figure 9.12 of \citet{IPCC} uses a regular gridded time series over a spatial domain to illustrate temporal patterns in climate change data.) It hides the irregular nature of the locations, though, and a complication is that icons may end up centered over nonsense places, such as land measurements in the ocean.

@@

Glyphs, of course, can be more general than lines or stars. For multivariate data we may want to represent several variables ($z_1, \dots, z_p; p=$ number of variables) in the glyphs. Figure~\ref{fig:cloud} shows scatterplots of the EXPO data as icons: temperature values ($z_1$ instead of $t$) are displayed horizontally and high-cloud values ($z_2$) are displayed vertically, both locally scaled. The bivariate relationship between the two variables is explored spatially: the equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association. The right side plot displays a loess curve fit to the data instead of the individual points, sharpening the association signal realtive to the variation. Gauguin \citep{gribov:2006}, has an icon based on a histogram, giving small univariate distribution displays for each location. \citet{carr:1992} use ray glyphs to show bivariate relationships.

@@

There are two natural next directions for this work: perceptual experiments, and development of interactive graphics. Several careful studies comparing the perceptual elements of these displays, using different scale and glyph options, and in comparison with facetted heatmaps, would be recommended. For example, it appears that periodic trends are easier to perceive in star-glyphs, and long-term trends in line-glyphs. Is this always the case? Are their certain types of periodicity that are particularly easy to spot? Rigorous perceptual studies will help guide the use of glyph-maps.

The conceptualization and equations for computing the glyph-maps lend themselves to producing these displays interactively. Different variables can be swept into the display using mouse motion mapped to $\alpha_x, \alpha_y$ (Equation \ref{coords.eqn}), resolution could be increased, and aggregated data icons disengaged, upon zooming into a spatial neighborhood. This experimentation is ideally done using R, so that computation can be linked with displays. The hurdle has been the lagging development of interactive graphics in R. Several new packages  in R have become available in the past year \citep{qtbase, qtpaint, plumbr} that support interactive graphics which will enable these developments.

In summary, glyph-maps enable the study of temporal patterns in multivariate spatiotemporal data. Climate change is focused on changes over time, so glyph-maps provide a way to explore these changes directly. Glyph-maps enable different resolutions of the data to be examined: the raw data to discover data quality issues, global trend, seasonality, residuals, and multivariate dependencies. When the data is spatially gridded the glyphs are organized in a manner that makes reasonable comparisons. For irregularly gridded data the icon size needs to be chosen to minimize overlap but use as much display space as possible, or combined to produce close to gridded icons.

\section*{Acknowledgements}

This work was partially supported National Science Research grant DMS0706949.
